---
title: "EDS 222: Homework 1"
date: "10/14/2024"
author: "Matteo Torres"
execute: 
  eval: true
  message: false
  warning: false
  output: false
format:
  html:
    toc: true
editor_options:
  chunk_output_type: console
---

## Background

*(The case study in this exercise is based on reality, but does not include actual observational data.)*

In this exercise we will look at a case study concerning air quality in South Asia. The World Health Organization estimates that air pollution kills an estimated seven million people per year, due to its effects on the cardiovascular and respiratory systems. Out of the 40 most polluted cities in the world, South Asia is home to 37, and Pakistan was ranked to contain the second most air pollution in the world in 2020 (IQAIR, 2020). In 2019, Lahore, Pakistan was the 12th most polluted city in the world, exposing a population of 11.1 million people to increased mortality and morbidity risks.

In this exercise, you are given two datasets from Lahore, Pakistan and are asked to compare the two different data collection strategies from this city. These data are:

-   Crowd-sourced data from air quality monitors located in people's homes. These data are voluntarily collected by individual households who choose to install a monitor in their home and upload their data for public access.

-   Official government data from monitors installed by government officials at selected locations across Lahore. There have been reports that government officials strategically locate monitors in locations with cleaner air in order to mitigate domestic and international pressure to clean up the air.

::: callout-note
All data for EDS 222 will be stored on the Taylor server, in the shared `/courses/eds-222/data/` directory. Please see material from EDS 214 on how to access and retrieve data from Taylor. These data are small; all compute can be handled locally. Thanks to Bren PhD student Fatiq Nadeem for assembling these data!
:::

In answering the following questions, please consider the lecture content from class on sampling strategies, as well as the material in Chapter 2 of [*Introduction to Modern Statistics*](https://openintro-ims.netlify.app/data-design). Include in your submission your version of this file "`eds-222-hw1.qmd`" and the rendered HTML output, each containing complete answers to all questions *as well as the associated code*. Questions with answers unsupported by the code will be marked incomplete. Showing your work this way will help you develop the habit of creating reproducible code.

## Assessment

```{r}
# Load packages
library(tidyverse)
```

### Question 1

Load the data from each source and label it as `crowdsourced` and `govt` accordingly. For example:

``` r
crowdsourced <- readRDS(file.path("data", "airpol-PK-crowdsourced.RDS"))
govt <- readRDS(file.path("data", "airpol-PK-govt.RDS"))
```
::: callout-warning
There's an implicit assumption about file organization in the code above. What is it? How can you make the code work?
:::

I successfully made the file path work from the root directory by starting at the "HW1" data folder then calling the file name I want to assign to each variable.

```{r}
# Read in data
crowdsourced <- readRDS(file.path("HW1", "airpol-PK-crowdsourced.RDS"))
govt <- readRDS(file.path("HW1", "airpol-PK-govt.RDS"))
```

1.  These dataframes have one row per pollution observation. How many pollution records are in each dataset?
```{r}
# Check number of rows/columns
dim(crowdsourced) # 5488 observations
dim(govt) # 1960 observations
```

The crowdsourced dataset has 5488 pollution records and the government dataset has 1960 pollution records.

2.  Each monitor is located at a unique latitude and longitude location. How many unique monitors are in each dataset?

::: callout-tip
`group_by(longitude,latitude)` and `cur_group_id()` in `dplyr` will help in creating a unique identifier for each (longitude, latitude) pair.
:::

```{r}
# Create a column for monitors in crowdsourced df
cs_monitors <- crowdsourced %>%
  group_by(longitude, latitude) %>%
  mutate(monitors = cur_group_id())

# Count unique monitors
length(unique(cs_monitors$monitors)) # 14 monitors
```

```{r}
# Create a column for monitors in govt df
g_monitors <- govt %>%
  group_by(longitude, latitude) %>%
  mutate(monitors = cur_group_id())

# Count unique monitors
length(unique(g_monitors$monitors)) # 5 monitors
```

The crowdsourced dataset has 14 unique monitors and the government dataset has 5 unique monitors.

### Question 2

The goal of pollution monitoring in Lahore is to measure the average pollution conditions across the city.

1.  What is the *population* in this setting? Please be precise.

The population in this study is the air quality in all of Lahore, Pakistan between November 2018 and November 2019.

2.  What are the *samples* in this setting? Please be precise.

The samples in this study are both the crowdsourced and government sourced datasets.

3.  These samples were not randomly collected from across locations in Lahore. Given the sampling approaches described above, discuss possible biases that may enter when we use these samples to construct estimates of population parameters.

One potential bias is volunteer bias, as the participants who choose to take part in the study might be limited or concentrated in specific locations in the city. This can negatively impact the results of the study because it may not be a representative example of the population. Additionally, there might be a selection bias for government monitors, as the chosen areas were selected based on their cleaner air quality standards rather than through random selection. This could impact the accuracy of the PM measurements collected for the study.

### Question 3

1.  For both the government data and the crowd-sourced data, report the sample mean, sample minimum, and sample maximum value of PM 2.5 (measured in $\mu g/m^3$).

```{r}
# Sample mean, min, and max of PM for crowdsourced data
mean(crowdsourced$PM) # 70.2008
min(crowdsourced$PM) # 20
max(crowdsourced$PM) # 120
```
The sample mean is 70.2 $\mu g/m^3$.

The sample minimum is 20 $\mu g/m^3$.

The sample maximum is 120 $\mu g/m^3$.


```{r}
# Sample mean, min, and max of PM for govt data
mean(govt$PM) # 39.64694
min(govt$PM) # 15
max(govt$PM) # 65
```
The sample mean is 39.6 $\mu g/m^3$.

The sample minimum is 15 $\mu g/m^3$.

The sample maximum is 65 $\mu g/m^3$.

2.  Discuss any key differences that you see between these two samples.

Both datasets have similar minimum values; however, the maximum value from the crowdsourced dataset is nearly double the maximum value from the government dataset. The significantly higher mean value also supports the idea that, on average, most crowdsourced data points are higher than the maximum value of the government dataset.

3.  Are the differences in mean pollution as expected, given what we know about the sampling strategies?

I believe the differences in mean pollution are as expected based on the sampling strategies because government monitoring was intended to have significantly lower PM readings than the crowdsourced monitoring to mitigate domestic and international pressures.

### Question 4

Use the location of the air pollution stations for both of the sampling strategies to generate a map showing locations of each observation. Color the two samples with different colors to highlight how each sample obtains measurements from different parts of the city.

::: callout-tip
`longitude` indicates location in the *x*-direction, while `latitude` indicates location in the *y*-direction. With `ggplot2` this should be nothing fancy. We'll do more spatial data in `R` later in the course.
:::

```{r, output = TRUE}
# Define colors for plot
legend_colors <- c("crowdsourced" = "red", "government" = "blue")

# Make plot of air pollution stations
stations <- ggplot() +
  geom_point(data = crowdsourced, aes(x = longitude, y = latitude, color = "crowdsourced")) +
  geom_point(data = govt, aes(x = longitude, y = latitude, color = "government")) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "Air Pollution Stations in Lahore, Pakistan",
       color = "Legend") +
  scale_color_manual(values = legend_colors) +
  theme_bw()

# Display plot
stations
```

### Question 5

The local newspaper in Pakistan, *Dawn*, claims that the government is misreporting the air pollution levels in Lahore. Do the locations of monitors in question 4, relative to crowd-sourced monitors, suggest anything about a possible political bias?

Yes, the government monitors are concentrated in a small area of Lahore, Pakistan. They purposefully do not portray average PM measurements across the entire city. This suggests that there is political bias because the government is censoring significant information, air quality data, from the public that could potentially help improve the quality of life. This minimizes the issues of pollution and public health to make politicians/leaders look more palatable.

### Question 6

Given the recent corruption in air quality reporting, the Prime Minister of Pakistan has hired an independent body of environmental data scientists to create an unbiased estimate of the mean PM 2.5 across Lahore using some combination of both government stations and crowd sourced observations.

NASA's satellite data indicates that the average PM across Lahore is 89.2 $\mu g/m^3$. Since this is the most objective estimate of population-level PM 2.5 available, your goal is to match this mean as closely as possible by creating a new ground-level monitoring sample that draws on both the government and crowd-sourced samples.

#### Question 6.1

First, generate a *random sample* of size $n=1000$ air pollution records by (i) pooling observations across the government and the crowd-sourced data; and (ii) drawing observations at random from this pooled sample.

::: callout-tip
`bind_rows()` may be helpful.
:::

```{r}
# Combine both crowdsourced and govt datasets
combined <- bind_rows(crowdsourced, govt)

# Check number of rows/columns
dim(combined) # 7448 observations

# Collect a random sample of 1000 observations from the combined dataset
samp_1000 <- combined %>%
  sample_n(1000)

# Verify sample size
dim(samp_1000)
```

Second, create a *stratified random sample*. Do so by (i) stratifying your pooled data-set into strata of 0.01 degrees of latitude, and (ii) randomly sampling 200 air pollution observations from each stratum.

```{r}
# Create a new column with stratified data (0.01 degrees lat)
combined <- combined %>%
  mutate(strat_2 = round(latitude, 2))

# Verify new column
dim(combined) # 5 columns

# Collect a random sample of 200 observations from each stratum
samp_200 <- combined %>%
  group_by(strat_2) %>%
  sample_n(200) %>%
  ungroup()

# Verify sample size
dim(samp_200) # 1000 (200 from each group)
```

#### Question 6.2

Compare estimated means of PM 2.5 for each sampling strategy to the NASA estimate of 89.2 $\mu g/m^3$. Which sample seems to match the satellite data best? What would you recommend the Prime Minister do? Does your proposed sampling strategy rely more on government or on crowd-sourced data? Why might that be the case?

```{r}
# Mean PM 2.5 for random
mean(samp_1000$PM)

# Mean PM 2.5 for stratified
mean(samp_200$PM)
```

random sample mean < stratified sample mean

The stratified random sample seems to match the NASA satellite data best. I recommend that the Prime Minister advocates for the people by expanding government monitors and endorsing the use of the stratified dataset for reporting.

```{r}
# monitors of the stratified data
monitors_samp <- samp_200 %>%
  group_by(longitude, latitude) %>%
  mutate(monitors = cur_group_id())

# number of observations per monitor
monitors_samp %>%
  group_by(monitors) %>%
  summarize(n())

# government monitors from stratified data based on lat/long
monitors_samp %>%
  filter(monitors %in% c(2, 3, 5, 6, 7)) %>%
  nrow() # observations from govt data

# crowdsourced monitors from stratified data based on lat/long
monitors_samp %>% 
  filter(monitors %in% c(1, 4, 8:19)) %>%
  nrow() # observations from crowdsourced data
```

As you can see above, the stratified data relies more on crowdsourced data. The purpose of the strata is to reduce the influence of government monitors on assessing air quality for the entire city, as they only cover a small localized portion. This approach provides the opportunity to sample and study other areas of Lahore. 


